### 2016-06-18 paper/icfp16-cameraready/abstract.tex
mention user study in abstract and intro

\begin{abstract}
  %
  \emph{Static} type errors are a common stumbling block
  for newcomers to typed functional languages.
  %
  We present a \emph{dynamic} approach to explaining type
  errors by generating counterexample witness inputs that
  illustrate \emph{how} an ill-typed program goes wrong.
  %
  First, given an ill-typed function, we symbolically
  execute the body to dynamically synthesize witness values
  that can make the program go wrong.
  We prove that our procedure synthesizes
  \emph{general witnesses} in that if a witness is
  found, then \emph{for all} inhabited input types, there exist
  values that can make the function go wrong.
  %
  Second, we show how to extend the above procedure to
  produce a \emph{reduction graph} that can be used to
  interactively visualize and debug witness executions.
  %
  Third, we evaluate our approach on two data sets
  comprising over 4,500 ill-typed student programs.
  Our technique is able to generate witnesses for
  88\% of the programs, and our reduction graph
  yields small counterexamples for 81\% of the witnesses.
\end{abstract}

>>>>

\begin{abstract}
  %
  \emph{Static} type errors are a common stumbling block
  for newcomers to typed functional languages.
  %
  We present a \emph{dynamic} approach to explaining type
  errors by generating counterexample witness inputs that
  illustrate \emph{how} an ill-typed program goes wrong.
  %
  First, given an ill-typed function, we symbolically
  execute the body to dynamically synthesize witness values
  that can make the program go wrong.
  We prove that our procedure synthesizes
  \emph{general witnesses} in that if a witness is
  found, then \emph{for all} inhabited input types, there exist
  values that can make the function go wrong.
  %
  Second, we show how to extend the above procedure to
  produce a \emph{reduction graph} that can be used to
  interactively visualize and debug witness executions.
  %
  Third, we evaluate the coverage of our approach on two data sets
  comprising over 4,500 ill-typed student programs.
  Our technique is able to generate witnesses for
  88\% of the programs, and our reduction graph
  yields small counterexamples for 81\% of the witnesses.
  %
  Finally, we evaluate the utility of our witnesses in
  helping students understand and fix type errors, and
  find that students presented with our witnesses
  consistently show a greater understanding of type errors
  than those presented with a standard error message.
\end{abstract}

### 2016-06-18 paper/icfp16-cameraready/introduction.tex
mention user study in abstract and intro

Type errors are a common stumbling block for students
trying to learn typed functional languages like \ocaml\
and \haskell.
%
Consider the ill-typed @fac@ function on the left in
Figure~\ref{fig:factorial}.
%
The function returns @true@ in the base case (instead of @1@),
and so \ocaml responds with the error message:
%
\begin{verbatim}
 File "fac.ml", line 5, characters 8-17:
 Error: This expression has type bool
        but an expression was expected
        of type int
\end{verbatim}
%
This message makes perfect sense to an expert who is familiar
with the language and has a good mental model of how the type
system works.
%
However, it may perplex a novice who has yet to develop such a
mental model.
%
To make matters worse, unification-based type inference algorithms
often report errors far removed from their source.
%
This further increases the novice's confusion and can actively mislead
them to focus their investigation on an irrelevant piece of code.
%
Much recent work has focused on analyzing unification constraints
to properly \emph{localize} a type error~\cite{Lerner2007-dt,Chen2014-gd,Zhang2014-lv,Pavlinovic2014-mr},
but an accurate source location still does not explain \emph{why} the
program is wrong.

@@

\paragraph{1. Finding Witnesses}
Our first contribution is an algorithm for searching for
\emph{witnesses} to type errors, \ie\ inputs that cause a
program to go wrong~(\S~\ref{sec:searching-witness}).
%
This problem is surprisingly tricky when we cannot rely on
static type information.
%
In particular, we must avoid the trap of \emph{spurious} inputs
that cause irrelevant problems that would be avoided by picking
values of a different, relevant type.
%
We solve this problem by developing a novel operational semantics
that combines evaluation and type inference.
%
We execute the program with \emph{holes} --- values whose type is
unknown --- as the inputs.
% that corresponds to executing the program with \emph{holes} ---
% values whose type is unknown.
%
A hole remains abstract until the evaluation context tells us what type
it must have, for example the parameters to an addition operation must
both be integers.
%
Our semantics \emph{conservatively} instantiates holes with concrete
values; thereby dynamically inferring the type of the input
until the program goes wrong.
%
We prove that our procedure synthesizes \emph{general witnesses}, which
means, intuitively, that if a witness is found for a given ill-typed
function, then, \emph{for all} (inhabited) input types, there exist
values that can make the function go wrong.

@@

We introduce a notion of \emph{jump-compressed} traces to abstract away
the irrelevant details of a computation.
%
A jump-compressed trace includes only big steps to the next function
call or return, for example the trace in the bottom-left of
Figure~\ref{fig:factorial} is a jump-compressed trace.
%
Jump-compressed traces are similar to stack traces in that both show a
sequence of function calls that lead to a crash, but the jump-compressed
trace also shows the return values of successful calls, which can be
useful in understanding why a particular path was taken.

@@

\smallskip
Our results show that in the vast majority of
cases, (novices') ill-typed programs \emph{do} go wrong.
This, in turn, opens the door to a novel dynamic way to
explain, understand, and appreciate the benefits of static typing.

>>>>

Type errors are a common stumbling block for students
trying to learn typed functional languages like \ocaml\
and \haskell.
%
Consider the ill-typed @fac@ function on the left in
Figure~\ref{fig:factorial}.
%
The function returns @true@ in the base case (instead of @1@),
and so \ocaml responds with the error message:
%
\begin{verbatim}
 File "fac.ml", line 5, characters 8-17:
 Error: This expression has type bool but an
        expression was expected of type int.
\end{verbatim}
%
This message makes perfect sense to an expert who is familiar
with the language and has a good mental model of how the type
system works.
%
However, it may perplex a novice who has yet to develop such a
mental model.
%
To make matters worse, unification-based type inference algorithms
often report errors far removed from their source.
%
This further increases the novice's confusion and can actively mislead
them to focus their investigation on an irrelevant piece of code.
%
Much recent work has focused on analyzing unification constraints
to properly \emph{localize} a type error~\cite{Lerner2007-dt,Chen2014-gd,Zhang2014-lv,Pavlinovic2014-mr},
but an accurate source location does not explain \emph{why} the
program is wrong.

@@

\paragraph{1. Finding Witnesses}
Our first contribution is an algorithm for searching for
\emph{witnesses} to type errors, \ie\ inputs that cause a
program to go wrong~(\S~\ref{sec:searching-witness}).
%
This problem is surprisingly tricky when we cannot rely on
static type information.
%
In particular, we must avoid the trap of \emph{spurious} inputs
that cause irrelevant problems that would be avoided by picking
values of a different, relevant type.
%
We solve this problem by developing a novel operational semantics
that combines evaluation and type inference.
%
We execute the program with \emph{holes} --- values whose type is
unknown --- as the inputs.
% that corresponds to executing the program with \emph{holes} ---
% values whose type is unknown.
%
A hole remains abstract until the evaluation context tells us what type
it must have, for example the parameters to an addition operation must
both be integers.
%
Our semantics \emph{conservatively} instantiates holes with concrete
values, dynamically inferring the type of the input
until the program goes wrong.
%
We prove that our procedure synthesizes \emph{general witnesses}, which
means, intuitively, that if a witness is found for a given ill-typed
function, then, \emph{for all} (inhabited) input types, there exist
values that can make the function go wrong.

@@

We introduce a notion of \emph{jump-compressed} traces to abstract away
the irrelevant details of a computation.
%
A jump-compressed trace includes only function
calls and returns, for example the trace in the bottom-left of
Figure~\ref{fig:factorial} is jump-compressed.
%
Jump-compressed traces are similar to stack traces, both show a
sequence of function calls that lead to a crash, but the jump-compressed
trace also shows the return values of successful calls, which can be
useful in understanding why a particular path was taken.

@@

\subparagraph{Witness Utility}
%
Even if we can find small witnesses for the majority of type errors, it
may be that the witnesses do not actually help developers
\emph{understand} the errors.
%
In other words, perhaps the static error message is sufficient to
diagnose and fix the error, or perhaps the witness simply does not add
enough information to make a difference.
%
Thus, our final contribution is a user study that compares the utility
of our witnesses with that of the error messages provided by the \ocaml
compiler~(\S~\ref{sec:user-study}).
%
We show that students given one of our witnesses are consistently more
likely to correctly explain and fix a type error than those given
the standard error message produced by the \ocaml compiler.

\smallskip
Our results show that in the vast majority of
cases, (novices') ill-typed programs \emph{do} go wrong,
and that the witnesses to these errors can be helpful in
understanding the source of the error.
%
This, in turn, opens the door to a novel dynamic way to
explain, understand, and appreciate the benefits of static typing.

### 2016-06-18 paper/icfp16-cameraready/related.tex
mention user study in abstract and intro

In contrast to these approaches, we do not attempt to localize or fix
the type error. Instead we try to explain it to the novice user using a
dynamic witness that demonstrates how their program is not just
ill-typed but truly wrong. In addition, allowing users to run their
program (even knowing that it is wrong) enables experimentation and the
use of debuggers to step through the program and investigate its
evolution.

>>>>

In contrast to these approaches, we do not attempt to localize or fix
the type error. Instead we try to explain it to the user using a
dynamic witness that demonstrates how the program is not just
ill-typed but truly wrong. In addition, allowing users to run their
program % (even knowing that it is wrong)
enables experimentation and the
use of debuggers to step through the program and investigate its
evolution.
